---
title: "RTips"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intergrating R and C++

For my research I came accross a R challenge whcih involved multiple nested loops with group and subgroup structure. My problem involved updating a vector $x$ where for each group $x^{g}$, if $\|x^{g}\|_2> 4$ the group was set to zeros, otherwise each subgroup was updated as:
\begin{align}
x^{sg} = \|x^{sg}\|_2\times x^{sg} - x^{sg}
\end{align}
A simple R solution for this problem is:

```{r gsgUpdate1}
groups = rep(1:10^3, each = 20)
subgroups = rep(1:(10^3*5), each = 4)
n <- length(groups)
x <- rnorm(n)

# Update groups
updateGroups1 <- function(x, groups, subgroups){
  newx <- x # preallocate 
  ugroups <- unique(groups)
  for(i in ugroups){
    gind <- which(groups == i)
    if(norm(matrix(x[gind],nrow=1), type = "e") < 5){
      res <- rep(0, length(gind))
    } else{
      res <- updateSubgroup(x[gind],subgroups[gind])
    }
    newx[gind] <- res
  }
  return(newx)
}

# Update subgroup
updateSubgroup <- function(x, subgroups){
  usubgroups <- unique(subgroups)
  newx <- x
  for(i in usubgroups){
    ind <- which(subgroups == i)
    newx[ind] <- norm(matrix(x[ind], nrow = 1), type = "e")*x[ind] - x[ind]
  }
  return(newx)
}
```
In the application, the update can be repeated hundereds of times so speed is an issue especially in more realistic situations with much larger vectors.
```{r time1, echo=2}
library(microbenchmark)
s <- microbenchmark(updateGroups1(x, groups, subgroups), times = 10)
print(s)
```
## Using sapply
By using the apply wrappers I thought I'd see some speedup (and possible extension to milticore)
```{r sapplymeth}
sapplyGroups <- function(x,groups, subgroups){
  ugroups <- unique(groups)
  newx <- sapply(ugroups, function(i) subGroupFunction(x[which(groups == i)], subgroups[which(groups == i)]))
  return(as.numeric(newx))
}
subGroupFunction <- function(x, subgroups){
  if(norm(matrix(x,nrow=1), type = "e") < 5){
    res <- rep(0, length(x))
    } 
  else {
    usubgroups <- unique(subgroups)
    res <- sapply(usubgroups, function(j) subsubGroupFunction(x[which(subgroups == j)]))
  }
  return(as.numeric(res))
}
subsubGroupFunction <- function(x){
  return(norm(matrix(x, nrow = 1), type = "e")*x - x)
}
```

Comparing the codes gives some fustrating results:
```{r, echo=1}
s <- microbenchmark(updateGroups1(x, groups, subgroups), sapplyGroups(x,groups, subgroups), times = 10)
print(s)
```
This is because the sapply method is not preallocating variable size (so each itteration increases the size of the returned vector). Additional time is taken converting the list into a numeric object between sapplys. A good blog on the downfalls of memory allocation in R is http://www.r-bloggers.com/efficient-ragged-arrays-in-r-and-rcpp/.

## Cpp implementation
The Rcpp solution is surprisingly simple and much easier to implement than the sapply version. Some good starting points for including C++ in your R code are: Hadley Wickham's blog http://adv-r.had.co.nz/Rcpp.html and Dirk Eddelbuettel's blog http://dirk.eddelbuettel.com/code/rcpp.html. Addtionally the Armadillo library http://arma.sourceforge.net/docs.html  provides some great linear algebra functions within c++ that make programing easier. 

The folloing code was written in a seperate file from the previous R code. To compile the C++ funcitons the code is sourced and then every function with Rcpp::export is avaliable in the R environment. 
```{Rcpp, fastImp}
#include <RcppArmadillo.h>
// [[Rcpp::depends(RcppArmadillo)]]
using namespace Rcpp;
// C++ reads one line at a time so subfunctions are defined first
// Subgroup update
arma::vec subfun(arma::vec x, arma::vec subgroups){
  arma::vec res(x.size());
  arma::vec usubgroup = unique(subgroups);
  int nsgroups = usubgroup.size();
  
  for (int ka = 0; ka < nsgroups; ka++){
    arma::umat sgind = arma::find( subgroups == usubgroup(ka));
    res.elem(sgind) = x.elem(sgind)*arma::norm(x.elem(sgind),2) - x.elem(sgind);
  }
  return res;
}

// [[Rcpp::export]]
// Group update
arma::vec updateC(arma::vec x,arma::vec groups,arma::vec subgroups){
  int n = x.size();
  arma::vec res(n);
  arma::vec ugroups = unique(groups);
  int ngroups = ugroups.size();
  
  for (int ka = 0; ka < ngroups; ka++){
    arma::umat gind = arma::find( groups == ugroups(ka)); // find group indicies
    if(arma::norm(x.elem(gind)) < 5){
      res.elem(gind) = arma::zeros(gind.size());
    } else {
      res.elem(gind) = subfun(x.elem(gind), subgroups.elem(gind));
    }
  }
  return res;
}
```
Now to test the functions on a larger example:
```{r, echo=1:5}
groups = rep(1:10^4, each = 20)
subgroups = rep(1:(10^4*5), each = 4)
n <- length(groups)
x <- rnorm(n)
s <- microbenchmark(updateGroups1(x, groups, subgroups), sapplyGroups(x,groups, subgroups), updateC(x,groups, subgroups), times = 5)
print(s)
library(ggplot2)
autoplot(s)
```
